---
title: "Geodep"
author: "Clemence HUANG"
execute:
  echo: false
format:
  pdf:
    toc: true
editor: visual
---

```{r}
library(here)
library(dplyr)
library(readxl)
library(tidyr)
library(stringr)
library(purrr)
library(countrycode)
library(haven)
library(rnaturalearth)
library(sf)
library(knitr)


rm(list = ls())
here::i_am('geodep_shiny.Rproj')

DEP_YEAR <- 2024  
```

## 0. HS6 product descriptions

```{r}
hs6_description <- read_excel("1_Data/hs22_hs6.xlsx")
colnames(hs6_description)[2] <- "hs6"
```

## 1. Imports (GeoDep_M)

```{r}
imports_final <- read_dta("1_Data/geodep_M_hs22.dta") |>
  filter(year == DEP_YEAR) |>
  left_join(hs6_description, by = "hs6") |>
  select(-Classification,-`Parent Code`, -Level,-IsBasicLevel) |>
  mutate(sect_strategic = if_else(sect_crm == 1 | sect_dual_use == 1 | sect_health == 1 | sect_agrifood == 1 | sect_energy == 1, 1, 0)
    )
```

`GeoDep_M.dta` is one row per (importer, HS6): it carries the destination's total import value (`import_dpt`), the four dependency criteria (`c1_M` to `c4_M`), the `dependent` flag, the identity/share of the *leading* supplier (`first_odpt`, `share_odpt`), and the `sect_*` flags.

## 2. Exports (GeoDep_X)

```{r}
exports_final <- read_dta("1_Data/geodep_X_hs22.dta") |>
  filter(year == DEP_YEAR) |>
  left_join(hs6_description, by = "hs6") |>
  select(-Classification,-`Parent Code`, -Level,-IsBasicLevel)
```

`GeoDep_X.dta` is the symmetric export-side table: one row per (exporter, HS6), with `export_opt`, `c1_X` to `c4_X`, `dependent`, and the leading destination's identity/share (`first_dpto`, `share_dpto`). No sector columns.

## 3. Descriptive-stats precomputation

```{r}
eu_countries <- c("DEU", "AUT", "BEL", "BGR", "CYP", "HRV", "DNK", "ESP",
                  "EST", "FIN", "FRA", "GRC", "HUN", "IRL", "ITA", "LVA",
                  "LTU", "LUX", "MLT", "NLD", "POL", "PRT", "ROU", "SVK",
                  "SVN", "SWE", "CZE")

sector_names <- c(
  "sect_crm" = "Critical Raw Materials",
  "sect_dual_use" = "Dual Use",
  "sect_health" = "Health",
  "sect_agrifood" = "Agrifood",
  "sect_energy" = "Energy",
  "sect_other" = "Other"
)

sector_choices <- c(
  "All Sectors"             = "all",
  "Critical Raw Materials"  = "sect_crm",
  "Dual Use"                = "sect_dual_use",
  "Health"                  = "sect_health",
  "Agrifood"                = "sect_agrifood",
  "Energy"                  = "sect_energy",
  "Other"                   = "sect_other"
)

to_eun <- function(iso3) {
  if_else(iso3 %in% eu_countries, "EUN", iso3)
}

world_polygons_cache_path <- "1_Data/world_polygons_cache.rds"

world_polygons <- ne_countries(scale = "medium", returnclass = "sf") |>
  mutate(
    iso_a3 = case_when(
      !is.na(iso_a3)    & iso_a3    != "-99" & iso_a3    != "" ~ iso_a3,
      !is.na(iso_a3_eh) & iso_a3_eh != "-99" & iso_a3_eh != "" ~ iso_a3_eh,
      !is.na(adm0_a3)   & adm0_a3   != "-99" & adm0_a3   != "" ~ adm0_a3,
      TRUE ~ countrycode(name_long, origin = "country.name", destination = "iso3c")
    )
  ) |>
  filter(!is.na(iso_a3))

old_s2 <- sf::sf_use_s2()
sf::sf_use_s2(FALSE)

world_polygons <- world_polygons |>
  sf::st_wrap_dateline(options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180")) |>
  sf::st_make_valid()

sf::sf_use_s2(old_s2)

saveRDS(world_polygons, world_polygons_cache_path)

iso_name_lookup <- world_polygons |>
  sf::st_drop_geometry() |>
  group_by(iso_a3) |>
  slice(1) |>
  ungroup() |>
  select(iso_a3, name) |>
  bind_rows(tibble(iso_a3 = "EUN", name = "European Union"))

country_choices_lookup <- iso_name_lookup |>
  filter(!iso_a3 %in% eu_countries)

country_choices <- setNames(country_choices_lookup$iso_a3, country_choices_lookup$name)
country_choices <- country_choices[order(names(country_choices))]
country_choices_ui <- c("Select a country..." = "", country_choices)

dep_import_base <- imports_final |>
  filter(dependent == 1) |>
  mutate(
    iso_d   = to_eun(iso_d),
    iso_o   = to_eun(first_odpt),
    imports = import_dpt * share_odpt / 100
  ) |>
  select(iso_d, iso_o, hs6, imports, import_dpt,
         Description, starts_with("sect_"), first_odpt)

dep_export_base <- exports_final |>
  filter(dependent == 1) |>
  mutate(
    iso_o   = to_eun(iso_o),
    iso_d   = to_eun(first_dpto),
    imports = export_opt * share_dpto / 100
  ) |>
  select(iso_o, iso_d, hs6, Description, imports, export_opt, first_dpto)

traded_by_country <- imports_final |>
  mutate(iso_d = to_eun(iso_d)) |>
  select(iso_d, hs6, dependant_M_MC_t = dependent, c4_M_MC = c4_M,
         import_dpt, starts_with("sect_"))

traded_by_country_export <- exports_final |>
  mutate(iso_o = to_eun(iso_o)) |>
  select(iso_o, hs6, dependant_X_MC_t = dependent, c4_X_MC = c4_X,
         export_opt)

imports_sector_long_all <- imports_final |>
  mutate(iso_d = to_eun(iso_d)) |>
  mutate(dep_all = dependent) |>
  select(iso_d, hs6, dep_all, starts_with("sect_")) |>
  pivot_longer(cols = starts_with("sect_"), names_to = "Sector", values_to = "is_in_sector") |>
  filter(is_in_sector == 1) |>
  mutate(Sector_Name = sector_names[Sector])
```

```{r}

trade_isos <- c(
  imports_final$iso_d,
  imports_final$first_odpt,
  exports_final$iso_o,
  exports_final$first_dpto
) |>
  unique() |>
  na.omit() |>
  to_eun() |>
  unique()

country_choices_lookup <- iso_name_lookup |>
  filter(!iso_a3 %in% eu_countries) |>  
  filter(iso_a3 %in% trade_isos)          

country_choices <- setNames(country_choices_lookup$iso_a3, country_choices_lookup$name)
country_choices <- country_choices[order(names(country_choices))]
country_choices_ui <- c("Select a country..." = "", country_choices)
```

## 4. Save everything the app needs

```{r}
geodep_inputs <- list(
  DEP_YEAR                 = DEP_YEAR,
  eu_countries              = eu_countries,
  sector_names               = sector_names,
  sector_choices              = sector_choices,
  world_polygons                = world_polygons,
  iso_name_lookup                 = iso_name_lookup,
  country_choices                    = country_choices,
  country_choices_ui                   = country_choices_ui,
  dep_import_base                        = dep_import_base,
  dep_export_base                          = dep_export_base,
  traded_by_country                          = traded_by_country,
  traded_by_country_export                     = traded_by_country_export,
  imports_sector_long_all                        = imports_sector_long_all,
  imports_final = imports_final,
  exports_final = exports_final
)

saveRDS(geodep_inputs, "1_Data/geodep_shiny_inputs.rds")

cat("Precomputation done. Saved 1_Data/geodep_shiny_inputs.rds",
    "(", format(object.size(geodep_inputs), units = "MB"), ")\n")
```

